The error when inferring phylogenies with incipient species by a Birth-Death model

### 2018-04-12 article.tex
Method

This research's goal is to explore the effect of using an overly simplistic
BD prior on a PBD simulated reality. We provide a data set, 
that quantifies the inference error made in general, and explores
the effect of the way species trees are sampled from an incipient species tree.  
In brief, we simulate protracted phylogenies using the PBD process,
from which we sample a species tree. From that species tree, 
we simulate a DNA sequence alignment. Then, BEAST2 uses these alignments
to infer a posterior of phylogenies, using a BD prior. The difference
between the (BD) posterior phylogenies and (PBD) species tree is quantified.

@@

The parameter space the data set explores is inspired on 
[Etienne et al, 2014] for two reasons: (1) the academic
reason of being able to compare results, and (2) the practical
reason that the simulated phylogies will have enough taxa to
be informative, but not too big to be computationally overdemanding.
This research uses a complete superset of those parameter settings.
For the speciation initiation rate, we'll use $0.1$, $0.5$ and $1.0$ 
speciation initiation events per (good species) lineage per time unit, 
extending the set of
values used from one ($0.5$) to three. 
The speciation completion
rates used are $0.1$, $0.3$, $1.0$ and $10^6$ speciation completion
events per (incipient species) lineage per time unit, in which only
the last value is added as a computational proxy for infinity, which
makes the experiment fall back to a Birth-Death model (as speciation
is again instantaneous), allowing to measure the baseline error.
The extinction rates used are $0.0$, $0.1$, $0.2$ and $0.4$ 
extinction events per (good or incipient) lineage per time unit,
in which $0.4$ is a new extinction rate.
We investigate these parameters in a factorial fashion, except for
(1) those combinations in which the extinction rate equals or exceeds
the speciation initiation rate, and (2) the difference between
speciation initiation rate and extinction rate exceeds $0.8$. This is
done to prevent overly taxon-poor and taxon-rich phylogenies respectively.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \centering 
  \begin{tabular}{l l l}
    \hline
    Parameter             & Description & Values \\
    \hline
    \hline
    $b = b_g = b_i$       & Speciation initiation rate & 0.1, 0.5, 1.0 \\
    $\lambda$             & Speciation completion rate & 0.1, 0.3, 1.0, $10^6$ \\
    $\mu = \mu_g = \mu_i$ & Extinction rate & 0.0, 0.1, 0.2, 0.4 \\
    \hline
    $t_c$                 & Crown age & 15 \\
    $\sigma_c$            & Standard deviation around crown age & 0.001 \\
    $M$                   & Sampling method & Youngest for odd $R_i$, oldest for even $R_i$ \\
    $r$                   & Mutation rate & $\frac{1000}{15}$ \\
    $l_a$                 & DNA alignment length & $15K$ \\
    $f_i$                 & MCMC sampling interval & 1K or more \\
    $R_i$                 & RNG seed incipient tree & 1 to 40K \\
    $R_a$                 & RNG seed alignment simulation & $R_i$ \\
    $R_b$                 & RNG seed BEAST2 & $R_i$ \\
    \hline
  \end{tabular}
  \caption{
    Overview of the 12 simulation parameters. Above the horizontal line is 
    the biological parameter set.
  }
  \label{table:simulation_parameters}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

This research's goal is to explore the effect of using an overly simplistic
BD prior on PBD simulated phylogenies. We provide a data set, 
that quantifies the inference error made in general, and explores
the effect of the way species trees are sampled from an incipient species tree.  
In brief, we simulate protracted phylogenies using the PBD process,
from which we sample a species tree. From that species tree, 
we simulate a DNA sequence alignment. Then, BEAST2 uses these alignments
to infer a posterior of phylogenies, using a BD prior. The difference
between the (BD) posterior phylogenies and (PBD) species tree is quantified.

@@

The PBD model has five biological parameters (see \ref{table:parameters}), 
which we explore in a factorial fashion, excluding some combinations. 
We assume that the speciation initiation and extinction rates
of an incipient and good species are equal ($b_i = b_g$ and $\mu_i = \mu_g$),
as this will result in monophyletic incipient species trees.
We only simulate a PBD process for phylogenies in which
speciation initiation exceeds extinction rate ($b > \mu$),
and in which their difference is not too big ($b - \mu < 0.8$), 
to prevent overly taxon-poor and taxon-rich phylogenies respectively.
The parameter values chosen are a 
superset of \cite{etienne2014}, as these parameters result in reasonably
sized phylogenies and allows us to compare results. 
For the speciation initiation rate $b$, we'll use $0.1$, $0.5$ and $1.0$ 
speciation initiation events per (good species) lineage per time unit.
The speciation completion rates used are $0.1$, $0.3$, $1.0$ and $\infty$ speciation completion
events per (incipient species) lineage per time unit. For $\lambda = \infty$, 
the PBD model equals a BD model, which allows us to measure the baseline error.
The extinction rates used are $0.0$, $0.1$, $0.2$ and $0.4$ 
extinction events per (good or incipient) lineage per time unit.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \centering 
  \begin{tabular}{l l l}
    \hline
    Parameter             & Description & Values \\
    \hline
    \hline
    $b = b_g = b_i$       & Speciation initiation rate & 0.1, 0.5, 1.0 \\
    $\lambda$             & Speciation completion rate & 0.1, 0.3, 1.0, $10^6$ \\
    $\mu = \mu_g = \mu_i$ & Extinction rate & 0.0, 0.1, 0.2, 0.4 \\
    \hline
    $t_c$                 & Crown age & 15 \\
    $\sigma_c$            & Standard deviation around crown age & 0.001 \\
    $M$                   & Sampling method & Youngest for odd $R_i$, oldest for even $R_i$ \\
    $r$                   & Mutation rate & $\frac{1000}{15}$ \\
    $l_a$                 & DNA alignment length & $15K$ \\
    $f_i$                 & MCMC sampling interval & 1K or more \\
    $R_i$                 & RNG seed incipient tree & 1 to 40K \\
    $R_a$                 & RNG seed alignment simulation & $R_i$ \\
    $R_b$                 & RNG seed BEAST2 & $R_i$ \\
    \hline
  \end{tabular}
  \caption{
    Overview of the 12 simulation parameters. Above the horizontal line is 
    the biological parameter set.
  }
  \label{table:parameters}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
